Data Exploration

### 2019-01-29 src/data_exploration.Rmd
fuckit

# Package dependencies
You can load them using the following code which uses a function called [ipak](https://gist.github.com/stevenworthington/3178163). 
Note this function checks to see if the packages are installed first.
The "include=FALSE" supresses the package installation text appearing in the document...
```{r pre-install packages, include=FALSE}
packages <- c("plyr", "graphics", "ggplot2", "corrplot", "FSA", "car", "rworldmap") 
source("../src/ipak.R")
ipak(packages)
```

@@

ggplot(nafo0a, aes(x = salinity_depth, colour = nafo_zone)) + geom_density(na.rm = TRUE) + geom_density(data=nafo0b, na.rm = TRUE) + geom_density(data=nafo2g, na.rm = TRUE) + geom_density(data=nafo2h, na.rm = TRUE) + geom_density(data=nafo2j, na.rm = TRUE) + geom_density(data=nafo3k, na.rm = TRUE) + geom_density(data=nafo3l, na.rm = TRUE) + geom_density(data=nafo3n, na.rm = TRUE) + geom_density(data=nafo3o, na.rm = TRUE) + geom_density(data=nafo3ps, na.rm = TRUE) + geom_density(data=nafo4r, na.rm = TRUE) + geom_density(data=nafo4s, na.rm = TRUE) + geom_density(data=nafo4t, na.rm = TRUE) + geom_density(data=nafo4vn, na.rm = TRUE) + geom_density(data=nafo4vs, na.rm = TRUE) + geom_density(data=nafo4w, na.rm = TRUE) + geom_density(data=nafo4x, na.rm = TRUE) + geom_density(data=nafo5y, na.rm = TRUE) + geom_density(data=nafohudson, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_nafo_plots/o2_depth_nafo.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

@@

@@

this needs to be 3D so you need to create a cell_id_3d
```{r}
presab$cell_id_3d <- paste(presab$cell_id, presab$depthlayerno, sep="_") #create a new column that is a unique cell_id & depth ID. 
head(presab)
```
and now do one with a timeslice label
```{r}
presab$cell_id_xyzt <- paste(presab$cell_id_3d, presab$year, presab$month, sep="_") #create a new column that is a unique cell_id & depth ID. 
head(presab)
```

the data is in presab$total_cell_obs_xyt 
```{r}
unique_pointpercell <- unique(presab$total_cell_obs_xyzt)
unique_pointpercell
```
bugger... ok so lets do a frequency table
```{r}
celobsmore1 <- subset(presab, total_cell_obs_xyzt > 1)
pointpercell <- with(celobsmore1, table(cell_id_xyzt, total_cell_obs_xyzt))
write.csv(pointpercell, file = "../output/bio/pointpercell.csv")
nrow(pointpercell)
pointpercell
```

so there are 949 3d timesliced cells that contain more than one occurrence. 
- different sampling days
- different coordinates
- risk of duplicate entries (but only if two instituions inputted the same data with slightly different coordinates and different reserch program names - see cell3985_18_2000_7 below as example)

@@

>>>>

# Package dependencies
You can load them using the following code which uses a function called [ipak](https://gist.github.com/stevenworthington/3178163). 
Note this function checks to see if the packages are installed first.
The "include=FALSE" supresses the package installation text appearing in the document...
```{r pre-install packages, include=FALSE}
packages <- c("plyr", "graphics", "ggplot2", "corrplot", "FSA", "car", "rworldmap", "fastmatch") 
source("../src/ipak.R")
ipak(packages)
```

@@

ggplot(nafo0a, aes(x = o2_depth, colour = nafo_zone)) + geom_density(na.rm = TRUE) + geom_density(data=nafo0b, na.rm = TRUE) + geom_density(data=nafo2g, na.rm = TRUE) + geom_density(data=nafo2h, na.rm = TRUE) + geom_density(data=nafo2j, na.rm = TRUE) + geom_density(data=nafo3k, na.rm = TRUE) + geom_density(data=nafo3l, na.rm = TRUE) + geom_density(data=nafo3n, na.rm = TRUE) + geom_density(data=nafo3o, na.rm = TRUE) + geom_density(data=nafo3ps, na.rm = TRUE) + geom_density(data=nafo4r, na.rm = TRUE) + geom_density(data=nafo4s, na.rm = TRUE) + geom_density(data=nafo4t, na.rm = TRUE) + geom_density(data=nafo4vn, na.rm = TRUE) + geom_density(data=nafo4vs, na.rm = TRUE) + geom_density(data=nafo4w, na.rm = TRUE) + geom_density(data=nafo4x, na.rm = TRUE) + geom_density(data=nafo5y, na.rm = TRUE) + geom_density(data=nafohudson, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_nafo_plots/o2_depth_nafo.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

@@

What I want to see if if there is a marked difference between the env. correlates of the presence points between gear type used. This is also to try deal with detection bias between gear type (b/c the whole region is not uniformly sampled by the same gear type)

first create gear datasets

```{r presence by gear}
presence$gear <- as.character(presence$gear)
bottom_trawl <- subset(presence, gear == "bottom_trawl")
bottom_trawl_alfredo_03 <- subset(presence, gear == "bottom_trawl_alfredo_03")
bottom_trawl_campelen_14 <- subset(presence, gear == "bottom_trawl_campelen_14")
bottom_trawl_campelen_1800 <- subset(presence, gear == "bottom_trawl_campelen_1800")
bottom_trawl_campelen_21 <- subset(presence, gear == "bottom_trawl_campelen_21")
bottom_trawl_cosmos_2600 <- subset(presence, gear == "bottom_trawl_cosmos_2600")
bottom_trawl_western_IIA <- subset(presence, gear == "bottom_trawl_western_IIA")
unknown <- subset(presence, gear == "unknown")
vertical_plankton_tow <- subset(presence, gear == "vertical_plankton_tow")
```

plot by each variable, less 3m (2 samples) 1c (one sample) and 5ze (zero samples?!)
```{r presence by gear by variable}
ggplot(bottom_trawl, aes(x = temp_surface, colour = gear)) + geom_density(na.rm = TRUE) + geom_density(data=bottom_trawl_alfredo_03, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_14, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_1800, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_21, na.rm = TRUE) + geom_density(data=bottom_trawl_cosmos_2600, na.rm = TRUE) + geom_density(data=bottom_trawl_western_IIA, na.rm = TRUE) + geom_density(data=unknown, na.rm = TRUE) + geom_density(data=vertical_plankton_tow, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_gear_plots/temp_surface_gear.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

ggplot(bottom_trawl, aes(x = temp_depth, colour = gear)) + geom_density(na.rm = TRUE) + geom_density(data=bottom_trawl_alfredo_03, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_14, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_1800, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_21, na.rm = TRUE) + geom_density(data=bottom_trawl_cosmos_2600, na.rm = TRUE) + geom_density(data=bottom_trawl_western_IIA, na.rm = TRUE) + geom_density(data=unknown, na.rm = TRUE) + geom_density(data=vertical_plankton_tow, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_gear_plots/temp_depth_gear.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

ggplot(bottom_trawl, aes(x = chl_surface, colour = gear)) + geom_density(na.rm = TRUE) + geom_density(data=bottom_trawl_alfredo_03, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_14, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_1800, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_21, na.rm = TRUE) + geom_density(data=bottom_trawl_cosmos_2600, na.rm = TRUE) + geom_density(data=bottom_trawl_western_IIA, na.rm = TRUE) + geom_density(data=unknown, na.rm = TRUE) + geom_density(data=vertical_plankton_tow, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_gear_plots/chl_surface_gear.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

ggplot(bottom_trawl, aes(x = chl_depth, colour = gear))  + geom_density(na.rm = TRUE) + geom_density(data=bottom_trawl_alfredo_03, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_14, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_1800, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_21, na.rm = TRUE) + geom_density(data=bottom_trawl_cosmos_2600, na.rm = TRUE) + geom_density(data=bottom_trawl_western_IIA, na.rm = TRUE) + geom_density(data=unknown, na.rm = TRUE) + geom_density(data=vertical_plankton_tow, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_gear_plots/chl_depth_gear.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

ggplot(bottom_trawl, aes(x = salinity_surface, colour = gear)) + geom_density(na.rm = TRUE) + geom_density(data=bottom_trawl_alfredo_03, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_14, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_1800, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_21, na.rm = TRUE) + geom_density(data=bottom_trawl_cosmos_2600, na.rm = TRUE) + geom_density(data=bottom_trawl_western_IIA, na.rm = TRUE) + geom_density(data=unknown, na.rm = TRUE) + geom_density(data=vertical_plankton_tow, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_gear_plots/salinity_surface_gear.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

ggplot(bottom_trawl, aes(x = salinity_depth, colour = gear)) + geom_density(na.rm = TRUE) + geom_density(data=bottom_trawl_alfredo_03, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_14, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_1800, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_21, na.rm = TRUE) + geom_density(data=bottom_trawl_cosmos_2600, na.rm = TRUE) + geom_density(data=bottom_trawl_western_IIA, na.rm = TRUE) + geom_density(data=unknown, na.rm = TRUE) + geom_density(data=vertical_plankton_tow, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_gear_plots/salinity_depth_gear.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

ggplot(bottom_trawl, aes(x = o2_surface, colour = gear)) + geom_density(na.rm = TRUE) + geom_density(data=bottom_trawl_alfredo_03, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_14, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_1800, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_21, na.rm = TRUE) + geom_density(data=bottom_trawl_cosmos_2600, na.rm = TRUE) + geom_density(data=bottom_trawl_western_IIA, na.rm = TRUE) + geom_density(data=unknown, na.rm = TRUE) + geom_density(data=vertical_plankton_tow, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_gear_plots/o2_surface_gear.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

ggplot(bottom_trawl, aes(x = o2_depth, colour = gear)) + geom_density(na.rm = TRUE) + geom_density(data=bottom_trawl_alfredo_03, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_14, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_1800, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_21, na.rm = TRUE) + geom_density(data=bottom_trawl_cosmos_2600, na.rm = TRUE) + geom_density(data=bottom_trawl_western_IIA, na.rm = TRUE) + geom_density(data=unknown, na.rm = TRUE) + geom_density(data=vertical_plankton_tow, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_gear_plots/o2_depth_gear.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

ggplot(bottom_trawl, aes(x = mlp_surface, colour = gear)) + geom_density(na.rm = TRUE) + geom_density(data=bottom_trawl_alfredo_03, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_14, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_1800, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_21, na.rm = TRUE) + geom_density(data=bottom_trawl_cosmos_2600, na.rm = TRUE) + geom_density(data=bottom_trawl_western_IIA, na.rm = TRUE) + geom_density(data=unknown, na.rm = TRUE) + geom_density(data=vertical_plankton_tow, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_gear_plots/mlp_gear.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

ggplot(bottom_trawl, aes(x = ssh_surface, colour = gear)) + geom_density(na.rm = TRUE) + geom_density(data=bottom_trawl_alfredo_03, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_14, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_1800, na.rm = TRUE) + geom_density(data=bottom_trawl_campelen_21, na.rm = TRUE) + geom_density(data=bottom_trawl_cosmos_2600, na.rm = TRUE) + geom_density(data=bottom_trawl_western_IIA, na.rm = TRUE) + geom_density(data=unknown, na.rm = TRUE) + geom_density(data=vertical_plankton_tow, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_gear_plots/ssh_gear.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png
```

@@

this needs to be 3D so you need to create a cell_id_3d
```{r cell_id_3d}
presab$cell_id_3d <- paste(presab$cell_id, presab$depthlayerno, sep="_") #create a new column that is a unique cell_id & depth ID. 
head(presab)
```
and now do one with a timeslice label
```{r cell_id_xyzt}
presab$cell_id_xyzt <- paste(presab$cell_id_3d, presab$year, presab$month, sep="_") #create a new column that is a unique cell_id & depth ID. 
head(presab)
write.csv(presab, "../output/bio/presab_cellid_xyzt.csv", row.names = FALSE)
```

because background points can land in the same xyzt, subset the dataset to be only presences

```{r}
presencedup <- subset(presab, occurrence == "1")
nrow(presencedup)
```

the data is in presencedup$total_cell_obs_xyzt
```{r unique_pointpercell}
unique_pointpercell <- unique(presencedup$total_cell_obs_xyzt)
unique_pointpercell
```
bugger... ok so lets do a frequency table
```{r unique_pointpercell frequencies}
pointpercellcount <- count(celobsmore1, "total_cell_obs_xyzt")
pointpercellcount$no_cells <- pointpercellcount$freq / pointpercellcount$total_cell_obs_xyzt
write.csv(pointpercellcount, file = "../output/bio/pointpercellcount.csv")
pointpercellcount
```

so there are 949 3d timesliced cells that contain more than one occurrence, and 2042 rows. 
- different sampling days
- different coordinates
- risk of duplicate entries (but only if two instituions inputted the same data with slightly different coordinates and different reserch program names - see cell3985_18_2000_7 below as example)

@@

Ideally you only want one observation cell cell_id_xyzt. First see if any are missing depth data

find the column numbers that correspond to the env. data (use fastmatch package as is quicker)

```{r}
fmatch("chl_surface", names(celobsmore1)) #28
fmatch("chl_depth", names(celobsmore1)) #29
fmatch("mlp_surface", names(celobsmore1)) #30
fmatch("o2_surface", names(celobsmore1)) #31
fmatch("o2_depth", names(celobsmore1)) #32
fmatch("salinity_surface", names(celobsmore1)) #33
fmatch("salinity_depth", names(celobsmore1)) #34
fmatch("ssh_surface", names(celobsmore1)) #35
fmatch("temp_surface", names(celobsmore1)) #36
fmatch("temp_depth", names(celobsmore1)) #37
```

then subset any with NA values

```{r celobsmore1_novals}
celobsmore1_novals <-  celobsmore1[!complete.cases(celobsmore1[28:37]), ] #28:37 represent the columns w/ env. data
nrow(celobsmore1_novals) #just to check how many there are - 1095 out of 2042
```

ok and subset with all values

```{r celobsmore1_allval}
celobsmore1_allvals <- celobsmore1[complete.cases(celobsmore1[28:37]), ] #28:37 represent the columns w/ env. data
nrow(celobsmore1_allvals) #947 out of 2042
```

ok so how many unique cells do we have with complete data...

```{r}
pointpercellcount_alldata <- count(celobsmore1_allvals, "total_cell_obs_xyzt")
pointpercellcount_alldata$no_cells <- pointpercellcount_alldata$freq / pointpercellcount_alldata$total_cell_obs_xyzt
write.csv(pointpercellcount_alldata, file = "../output/bio/pointpercellcount_alldata.csv")
pointpercellcount_alldata
```

and see which cell_id_xyzt appear in both celobsmore1_allvals and celobsmore1_novals
```{r}
celobsmore1_allnovals <- inner_join(celobsmore1_allvals, celobsmore1_novals)
celobsmore1_allnovals 
```
None.... which is good. I'd expect observations in the same xyzt to have the same values

so, I can successfully reduce the observations down to one per xyzt

```{r}
fmatch("cell_id_xyzt", names(celobsmore1)) #46
duplicatesbegone <- celobsmore1[!duplicated(celobsmore1[ ,46]), ] #creates a df with all duplicates gone (now 1 obs per cell)
nrow(duplicatesbegone) #949 - good
presabnodup <- subset(presab, total_cell_obs_xyzt == "1") #remove all rows that have more than one obs
nrow(presabnodup) #1549351 - this adds up 
presab <- rbind(presabnodup, duplicatesbegone)
nrow(presab) #1550300 - great!

#recalculate the total_cell_obs_xyzt column

names(presab)[names(presab)=="total_cell_obs_xyzt"] <- "XXtotal_cell_obs_xyzt" #rename col
obs_cell_yymm_depth_dup <- count(presab, cell_id, year, month, depthlayerno)
test <- merge(presab, obs_cell_yymm_depth_dup, by.x=c("cell_id", "year", "month", "depthlayerno"), by.y=c("cell_id", "year", "month", "depthlayerno"))
names(test)[names(test)=="n"] <- "total_cell_obs_xyzt" #rename col

write.csv(presab, "../output/bio/presab_cellid_xyzt_nodup.csv")
rm(presabnodup, duplicatesbegone, celobsmore1, celobsmore1_novals, celobsmore1_allnovals, celobsmore1_allvals) #just because they are bigish variables
```
